---
title: "Employee Attrition: Binary Classification Lab"
subtitle: "From operational questions to retention decisions"
author: "Analytics Solutions Bootcamp"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: flatly
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5
)

required_packages <- c(
  "modeldata", "tidyverse", "janitor", "rsample",
  "rpart", "rpart.plot", "yardstick", "broom", "scales"
)

missing_packages <- required_packages[
  !required_packages %in% rownames(installed.packages())
]

if (length(missing_packages) > 0) {
  stop(
    "Install the missing packages first: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(modeldata)
library(tidyverse)
library(janitor)
library(rsample)
library(rpart)
library(rpart.plot)
library(yardstick)
library(broom)
library(scales)

set.seed(123)
```

# Lab scenario

An organization wants to use its historical employee records to support a
targeted retention program. The goal is not simply to build the model with the
highest accuracy. HR needs a solution that can identify employees who may need
support while keeping the number of employees contacted manageable.

The response variable is `attrition`:

- `Yes`: the employee left the organization.
- `No`: the employee stayed.

## Operational questions

This lab will answer the following questions:

1. What proportion of employees left the organization?
2. Which employee groups have relatively high historical attrition rates?
3. Can employee information help HR prioritize employees for a retention review?
4. Which model—logistic regression or a decision tree—better supports the
   operational goal?
5. How many actual attrition cases would each model miss?
6. If HR can only review a limited number of employees, how should the
   classification threshold be selected?
7. What patterns can HR investigate without incorrectly treating model results
   as causal explanations?

> **Important:** This is a synthetic dataset intended for instruction. A real
> employee-risk system requires privacy review, fairness testing, human
> oversight, and validation using current organizational data. Predictions
> should support employee assistance—not punitive decisions.

# 1. Load and understand the data

```{r load-data}
data(attrition, package = "modeldata")

attrition_data <- attrition |>
  clean_names() |>
  mutate(attrition = factor(attrition, levels = c("No", "Yes")))

glimpse(attrition_data)
```

```{r data-size}
tibble(
  employees = nrow(attrition_data),
  variables = ncol(attrition_data),
  duplicate_rows = sum(duplicated(attrition_data)),
  missing_values = sum(is.na(attrition_data))
)
```

Before modeling, confirm that the response has exactly two possible classes.

```{r target-check}
attrition_data |>
  count(attrition) |>
  mutate(proportion = percent(n / sum(n), accuracy = 0.1))
```

# 2. Explore the operational problem

## Overall attrition rate

```{r attrition-rate}
overall_attrition <- attrition_data |>
  summarise(
    employees = n(),
    employees_who_left = sum(attrition == "Yes"),
    attrition_rate = mean(attrition == "Yes")
  )

overall_attrition
```

```{r attrition-chart}
attrition_data |>
  count(attrition) |>
  mutate(share = n / sum(n)) |>
  ggplot(aes(attrition, n, fill = attrition)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(
    aes(label = paste0(n, " (", percent(share, accuracy = 0.1), ")")),
    vjust = -0.4
  ) +
  scale_fill_manual(values = c("No" = "#4C78A8", "Yes" = "#E45756")) +
  expand_limits(y = max(table(attrition_data$attrition)) * 1.1) +
  labs(
    title = "Historical employee attrition",
    subtitle = "The class distribution determines whether accuracy is sufficient",
    x = "Employee left the organization",
    y = "Number of employees"
  ) +
  theme_minimal()
```

## Where is attrition concentrated?

The following summaries help HR decide which workforce conditions deserve
further investigation. They describe associations and do not establish causes.

```{r operational-summary-function}
attrition_summary <- function(data, group_variable) {
  data |>
    group_by({{ group_variable }}) |>
    summarise(
      employees = n(),
      employees_who_left = sum(attrition == "Yes"),
      attrition_rate = mean(attrition == "Yes"),
      .groups = "drop"
    ) |>
    arrange(desc(attrition_rate))
}
```

```{r overtime-summary}
attrition_summary(attrition_data, over_time)
```

```{r job-role-summary}
attrition_summary(attrition_data, job_role)
```

```{r travel-summary}
attrition_summary(attrition_data, business_travel)
```

```{r satisfaction-summary}
attrition_summary(attrition_data, job_satisfaction)
```

```{r income-by-attrition}
attrition_data |>
  group_by(attrition) |>
  summarise(
    employees = n(),
    median_monthly_income = median(monthly_income),
    median_years_at_company = median(years_at_company),
    median_distance_from_home = median(distance_from_home),
    .groups = "drop"
  )
```

```{r overtime-chart}
attrition_summary(attrition_data, over_time) |>
  ggplot(aes(over_time, attrition_rate, fill = over_time)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(
    aes(label = percent(attrition_rate, accuracy = 0.1)),
    vjust = -0.4
  ) +
  scale_y_continuous(labels = percent, limits = c(0, NA)) +
  labs(
    title = "Historical attrition rate by overtime status",
    x = "Works overtime",
    y = "Attrition rate"
  ) +
  theme_minimal()
```

### Discussion

- Which groups have the highest attrition rates?
- Are these also large employee groups, or only small groups with high rates?
- What additional data would HR need before designing an intervention?
- Which variables might reflect a condition HR can improve?

# 3. Prepare training and test data

We remove administrative constants and identifiers because they do not provide
generalizable employee information. We also exclude `gender` from modeling to
avoid directly using a protected characteristic in individual risk scores.
In a real project, protected attributes may still be retained separately for
fairness auditing.

```{r modeling-data}
model_data <- attrition_data |>
  select(
    -any_of(c(
      "employee_number", "employee_count", "over18",
      "standard_hours", "gender"
    ))
  )

model_data |>
  summarise(across(everything(), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "missing") |>
  filter(missing > 0)
```

Attrition is the minority class, so we use a **stratified split** to preserve
approximately the same class proportions in the training and test sets.

```{r split-data}
attrition_split <- initial_split(
  model_data,
  prop = 0.80,
  strata = attrition
)

train_data <- training(attrition_split)
test_data  <- testing(attrition_split)

bind_rows(
  train_data |> count(attrition) |> mutate(dataset = "Training"),
  test_data  |> count(attrition) |> mutate(dataset = "Test")
) |>
  group_by(dataset) |>
  mutate(proportion = percent(n / sum(n), accuracy = 0.1)) |>
  ungroup()
```

The test set is kept untouched during model fitting. It represents employees
whose outcomes were not used to estimate either model.

# 4. Establish a frequency-based baseline

Following the **zero-rule** approach, the baseline predicts the majority class
(`No`) for every employee. A useful model should provide operational value
beyond this naive rule.

```{r baseline}
baseline_predictions <- test_data |>
  transmute(
    truth = attrition,
    .pred_class = factor("No", levels = levels(attrition))
  )

baseline_accuracy <- accuracy(
  baseline_predictions,
  truth = truth,
  estimate = .pred_class
)

baseline_accuracy
```

Accuracy may look high even though this baseline identifies no employees who
actually leave. This is why recall, precision, F1 score, and the confusion
matrix are necessary.

# 5. Build a logistic regression model

Logistic regression estimates each employee's probability of attrition. Its
coefficients can help identify associations after controlling for the other
included variables.

```{r logistic-model}
logistic_model <- glm(
  attrition ~ ., 
  data = train_data,
  family = binomial(link = "logit")
)

summary(logistic_model)
```

Odds ratios are easier to interpret than raw log-odds coefficients. Values
above 1 are associated with higher odds of attrition; values below 1 are
associated with lower odds, holding the other included predictors constant.

```{r logistic-odds-ratios}
logistic_odds_ratios <- tidy(
  logistic_model,
  conf.int = TRUE,
  exponentiate = TRUE
) |>
  filter(term != "(Intercept)") |>
  arrange(desc(abs(log(estimate))))

logistic_odds_ratios |>
  select(term, odds_ratio = estimate, conf.low, conf.high, p.value) |>
  slice_head(n = 15)
```

```{r logistic-predictions}
logistic_predictions <- test_data |>
  transmute(
    truth = attrition,
    .pred_yes = predict(
      logistic_model,
      newdata = test_data,
      type = "response"
    ),
    .pred_no = 1 - .pred_yes,
    .pred_class = factor(
      if_else(.pred_yes >= 0.50, "Yes", "No"),
      levels = c("No", "Yes")
    )
  )
```

# 6. Build a decision tree model

A decision tree creates a hierarchy of rules. Its main operational advantage is
that HR stakeholders can inspect the sequence of conditions leading to a
classification.

We set `cp` below its final value first, then use cross-validation error to
select and prune the tree.

```{r decision-tree-model}
tree_full <- rpart(
  attrition ~ ., 
  data = train_data,
  method = "class",
  parms = list(split = "information"),
  control = rpart.control(cp = 0.001, minsplit = 20, xval = 10)
)

printcp(tree_full)

best_cp <- tree_full$cptable[
  which.min(tree_full$cptable[, "xerror"]),
  "CP"
]

decision_tree <- prune(tree_full, cp = best_cp)
```

```{r plot-tree, fig.height=7}
rpart.plot(
  decision_tree,
  type = 2,
  extra = 104,
  fallen.leaves = TRUE,
  box.palette = c("#4C78A8", "#E45756"),
  shadow.col = "gray",
  nn = TRUE
)
```

```{r tree-importance}
tree_importance <- enframe(
  decision_tree$variable.importance,
  name = "variable",
  value = "importance"
) |>
  mutate(importance = importance / sum(importance)) |>
  arrange(desc(importance))

tree_importance
```

Variable importance identifies variables useful for splitting the historical
records. It does not prove that changing a variable will cause attrition to
decrease.

```{r tree-predictions}
tree_probabilities <- predict(
  decision_tree,
  newdata = test_data,
  type = "prob"
)

tree_predictions <- test_data |>
  transmute(
    truth = attrition,
    .pred_no = tree_probabilities[, "No"],
    .pred_yes = tree_probabilities[, "Yes"],
    .pred_class = factor(
      if_else(.pred_yes >= 0.50, "Yes", "No"),
      levels = c("No", "Yes")
    )
  )
```

# 7. Evaluate both models

For this retention use case, `Yes` is the event of interest. A false negative
is an employee who actually leaves but whom the model labels as likely to stay.
If early support is the objective, these missed cases may be more costly than
false positives.

```{r metric-function}
classification_metrics <- function(predictions, model_name) {
  bind_rows(
    accuracy(predictions, truth, .pred_class),
    precision(predictions, truth, .pred_class, event_level = "second"),
    recall(predictions, truth, .pred_class, event_level = "second"),
    f_meas(predictions, truth, .pred_class, event_level = "second"),
    roc_auc(predictions, truth, .pred_yes, event_level = "second")
  ) |>
    mutate(model = model_name, .before = 1)
}

model_metrics <- bind_rows(
  classification_metrics(logistic_predictions, "Logistic regression"),
  classification_metrics(tree_predictions, "Decision tree")
)

model_metrics |>
  select(model, metric = .metric, estimate = .estimate) |>
  pivot_wider(names_from = metric, values_from = estimate) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

## Confusion matrices

```{r logistic-confusion}
conf_mat(
  logistic_predictions,
  truth = truth,
  estimate = .pred_class
)
```

```{r tree-confusion}
conf_mat(
  tree_predictions,
  truth = truth,
  estimate = .pred_class
)
```

```{r operational-errors}
operational_counts <- function(predictions, model_name) {
  predictions |>
    summarise(
      model = model_name,
      employees_flagged = sum(.pred_class == "Yes"),
      correctly_flagged = sum(truth == "Yes" & .pred_class == "Yes"),
      unnecessary_reviews = sum(truth == "No" & .pred_class == "Yes"),
      missed_attrition_cases = sum(truth == "Yes" & .pred_class == "No")
    )
}

bind_rows(
  operational_counts(logistic_predictions, "Logistic regression"),
  operational_counts(tree_predictions, "Decision tree")
)
```

## ROC curves

```{r roc-curves}
bind_rows(
  logistic_predictions |> mutate(model = "Logistic regression"),
  tree_predictions |> mutate(model = "Decision tree")
) |>
  group_by(model) |>
  roc_curve(truth, .pred_yes, event_level = "second") |>
  autoplot() +
  geom_abline(linetype = "dashed", color = "gray50") +
  labs(
    title = "ROC curves on the test set",
    subtitle = "Performance across all possible classification thresholds",
    x = "False-positive rate",
    y = "True-positive rate",
    color = "Model"
  ) +
  theme_minimal()
```

# 8. Connect the threshold to HR capacity

The default threshold of 0.50 is a convention, not an operational requirement.
Lower thresholds generally flag more employees and improve recall, but they
also create more false positives and require more HR reviews.

```{r threshold-function}
threshold_results <- function(predictions, model_name,
                              thresholds = seq(0.10, 0.70, by = 0.05)) {
  map_dfr(thresholds, function(cutoff) {
    scored <- predictions |>
      mutate(
        threshold = cutoff,
        classification = factor(
          if_else(.pred_yes >= cutoff, "Yes", "No"),
          levels = c("No", "Yes")
        )
      )

    tibble(
      model = model_name,
      threshold = cutoff,
      employees_flagged = sum(scored$classification == "Yes"),
      recall = sens_vec(
        scored$truth,
        scored$classification,
        event_level = "second"
      ),
      precision = precision_vec(
        scored$truth,
        scored$classification,
        event_level = "second"
      ),
      missed_attrition_cases = sum(
        scored$truth == "Yes" & scored$classification == "No"
      )
    )
  })
}

threshold_comparison <- bind_rows(
  threshold_results(logistic_predictions, "Logistic regression"),
  threshold_results(tree_predictions, "Decision tree")
)

threshold_comparison
```

```{r threshold-chart}
threshold_comparison |>
  ggplot(aes(employees_flagged, recall, color = model)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  labs(
    title = "Retention-review workload versus recall",
    subtitle = "Each point represents a different classification threshold",
    x = "Employees flagged for HR review",
    y = "Share of actual attrition cases identified",
    color = "Model"
  ) +
  theme_minimal()
```

Suppose HR can review at most 20% of employees in the test population. The next
table finds the feasible threshold with the highest recall for each model.

```{r capacity-based-selection}
review_capacity <- ceiling(0.20 * nrow(test_data))

recommended_thresholds <- threshold_comparison |>
  filter(employees_flagged <= review_capacity) |>
  group_by(model) |>
  slice_max(recall, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(desc(recall), desc(precision))

tibble(review_capacity = review_capacity)
recommended_thresholds
```

An alternative production process is to rank employees by predicted
probability and review the top `r review_capacity` employees. This directly
matches workload to available capacity and avoids assuming that 0.50 is the
right cutoff.

```{r top-risk-review-list}
best_model_name <- recommended_thresholds$model[[1]]

selected_predictions <- if (best_model_name == "Logistic regression") {
  logistic_predictions
} else {
  tree_predictions
}

review_list_summary <- selected_predictions |>
  mutate(test_row = row_number()) |>
  slice_max(.pred_yes, n = review_capacity, with_ties = FALSE) |>
  summarise(
    selected_model = best_model_name,
    employees_for_review = n(),
    actual_attrition_cases_captured = sum(truth == "Yes"),
    precision_within_review_list = mean(truth == "Yes")
  )

review_list_summary
```

For privacy, the lab reports an aggregate summary rather than displaying an
employee-level watchlist.

# 9. Operational interpretation

Use the evidence generated above to answer these questions:

1. **Scale of the issue:** How many employees and what percentage of the
   workforce historically left?
2. **Workforce patterns:** Which roles or working conditions should HR study
   further? Are the affected groups large enough to prioritize?
3. **Baseline value:** Why is the majority-class accuracy an inadequate success
   measure for this use case?
4. **Model selection:** Which model identifies a larger share of actual
   attrition cases? Which one creates fewer unnecessary reviews?
5. **Capacity:** Under the 20% review constraint, which model and threshold
   should HR pilot?
6. **Interpretability:** Is the decision tree easier to communicate? Does that
   compensate for any difference in predictive performance?
7. **Intervention:** What supportive actions could HR test—for example workload
   discussions, career-path reviews, manager check-ins, or compensation review?
8. **Measurement:** How would HR determine whether an intervention actually
   reduces attrition rather than merely targeting correlated groups?

# 10. Recommended decision framework

The selected model should not automatically determine employment actions. A
safer operational workflow is:

1. Score eligible employees at an agreed schedule.
2. Select a threshold or ranked review list based on HR capacity.
3. Let trained HR personnel review the surrounding context.
4. Offer supportive, consistently defined interventions.
5. Track participation, employee feedback, retention outcomes, false alarms,
   and missed cases.
6. Monitor performance and fairness across relevant employee groups.
7. Retrain or retire the model when its data or operating environment changes.

# 11. Key limitations

- The data is synthetic and does not represent a specific organization.
- Historical associations are not proof of causation.
- Some relevant factors may be unavailable or poorly measured.
- A random split estimates generalization to similar records, not necessarily
  performance in a future workforce. Production validation should use a
  time-based holdout when dates are available.
- Using employee data creates privacy, fairness, explainability, and governance
  obligations.
- Model performance may change as workforce policies and economic conditions
  change.

# 12. Optional learner exercises

1. Change the review capacity from 20% to 10% and 30%. How does the recommended
   threshold change?
2. Compare the zero-rule baseline with a one-rule classifier based on
   `over_time`.
3. Modify the decision tree's `minsplit` or pruning rule. What happens to recall
   and interpretability?
4. Add 10-fold cross-validation to compare model stability.
5. Calculate specificity and balanced accuracy.
6. Examine model performance across gender groups using a separate audit table,
   without adding gender to the predictors.
7. Write a one-page recommendation to HR that distinguishes predictive
   associations from causal conclusions.

# Reproducibility information

```{r session-info}
sessionInfo()
```
