---
title: "Iris Species: Classification, Clustering, and PCA Lab"
subtitle: "Comparing supervised and unsupervised approaches to flower morphology"
author: "Analytics Solutions Bootcamp"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: flatly
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5
)

required_packages <- c(
  "tidyverse", "janitor", "rsample", "rpart", "rpart.plot",
  "yardstick", "cluster", "scales"
)

missing_packages <- required_packages[
  !required_packages %in% rownames(installed.packages())
]

if (length(missing_packages) > 0) {
  stop(
    "Install the missing packages first: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(tidyverse)
library(janitor)
library(rsample)
library(rpart)
library(rpart.plot)
library(yardstick)
library(cluster)
library(scales)

set.seed(123)
```

# Research scenario

A botanical field team studies three iris species using four measurements:
sepal length, sepal width, petal length, and petal width. The team wants to know
whether morphology can support species identification and whether natural
groups emerge even when species labels are hidden.

The lab compares four analytical approaches:

1. Decision tree using the original measurements
2. Best K-means model using an optimized subset of measurements
3. Decision tree using selected principal components
4. K-means using selected principal components

## Research questions

1. How do the four measurements vary among the three species?
2. Which combinations of measurements produce the most cohesive and separated
   natural clusters?
3. Does the optimal number of clusters agree with the three known species?
4. How closely do unlabeled clusters correspond to the actual species?
5. How many principal components are needed to retain at least 90% of the
   standardized measurement variance?
6. Does PCA preserve enough information for species identification and
   clustering?
7. Which of the four approaches best recovers the known species on held-out
   specimens?

> **Methodological note:** Species labels are used to train decision trees.
> K-means does not use species labels to form clusters. Feature combinations and
> cluster counts are selected using silhouette score on training measurements.
> Labels are consulted afterward only to evaluate and interpret the clusters.

# 1. Load and inspect the data

```{r load-data}
data(iris)

iris_data <- iris |>
  clean_names() |>
  mutate(species = factor(species))

feature_names <- c(
  "sepal_length", "sepal_width", "petal_length", "petal_width"
)

glimpse(iris_data)
```

```{r data-quality}
tibble(
  specimens = nrow(iris_data),
  variables = ncol(iris_data),
  duplicate_rows = sum(duplicated(iris_data)),
  missing_values = sum(is.na(iris_data))
)
```

```{r class-distribution}
iris_data |>
  count(species) |>
  mutate(proportion = percent(n / sum(n), accuracy = 0.1))
```

# 2. Explore flower morphology

```{r species-summary}
iris_data |>
  group_by(species) |>
  summarise(
    specimens = n(),
    across(
      all_of(feature_names),
      list(mean = mean, sd = sd),
      .names = "{.col}_{.fn}"
    ),
    .groups = "drop"
  )
```

```{r measurement-boxplots, fig.height=8}
iris_data |>
  pivot_longer(
    all_of(feature_names),
    names_to = "measurement",
    values_to = "centimeters"
  ) |>
  ggplot(aes(species, centimeters, fill = species)) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(~ measurement, scales = "free_y", ncol = 2) +
  labs(
    title = "Flower measurements by iris species",
    x = "Species",
    y = "Measurement (cm)"
  ) +
  theme_minimal()
```

```{r morphology-scatter}
ggplot(
  iris_data,
  aes(petal_length, petal_width, color = species, shape = species)
) +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    title = "Petal morphology of the three iris species",
    x = "Petal length (cm)",
    y = "Petal width (cm)",
    color = "Species",
    shape = "Species"
  ) +
  theme_minimal()
```

# 3. Create training and test samples

All optimization occurs using the training data. The test data is reserved for
the final four-model comparison.

```{r split-data}
iris_split <- initial_split(
  iris_data,
  prop = 0.80,
  strata = species
)

train_data <- training(iris_split)
test_data <- testing(iris_split)

bind_rows(
  train_data |> count(species) |> mutate(dataset = "Training"),
  test_data |> count(species) |> mutate(dataset = "Test")
) |>
  group_by(dataset) |>
  mutate(proportion = percent(n / sum(n), accuracy = 0.1)) |>
  ungroup()
```

# 4. Helper functions

The functions below keep preprocessing consistent and allow K-means clusters to
be applied to held-out specimens.

```{r helper-functions}
# Scale test data using means and standard deviations learned from training data.
standardize_pair <- function(train_matrix, test_matrix) {
  train_matrix <- as.matrix(train_matrix)
  test_matrix <- as.matrix(test_matrix)

  centers <- colMeans(train_matrix)
  scales <- apply(train_matrix, 2, sd)

  list(
    train = scale(train_matrix, center = centers, scale = scales),
    test = scale(test_matrix, center = centers, scale = scales),
    center = centers,
    scale = scales
  )
}

# Assign new rows to the nearest fitted K-means centroid.
nearest_cluster <- function(new_matrix, centers) {
  apply(new_matrix, 1, function(observation) {
    distances <- rowSums(
      (centers - matrix(
        observation,
        nrow = nrow(centers),
        ncol = ncol(centers),
        byrow = TRUE
      ))^2
    )
    which.min(distances)
  })
}

# Map each training cluster to its most frequent known species.
learn_cluster_labels <- function(clusters, truth) {
  tibble(cluster = clusters, species = truth) |>
    count(cluster, species) |>
    group_by(cluster) |>
    slice_max(n, n = 1, with_ties = FALSE) |>
    ungroup() |>
    select(cluster, species)
}

map_clusters_to_species <- function(clusters, label_map, species_levels) {
  lookup <- setNames(as.character(label_map$species), label_map$cluster)
  factor(lookup[as.character(clusters)], levels = species_levels)
}

# Adjusted Rand Index: label-invariant agreement between two partitions.
adjusted_rand_index <- function(truth, estimate) {
  tab <- table(truth, estimate)
  choose_2 <- function(x) x * (x - 1) / 2

  sum_cells <- sum(choose_2(tab))
  sum_rows <- sum(choose_2(rowSums(tab)))
  sum_cols <- sum(choose_2(colSums(tab)))
  total_pairs <- choose_2(sum(tab))
  expected <- sum_rows * sum_cols / total_pairs
  maximum <- (sum_rows + sum_cols) / 2

  (sum_cells - expected) / (maximum - expected)
}

macro_classification_metrics <- function(truth, estimate) {
  per_class <- map_dfr(levels(truth), function(current_class) {
    truth_binary <- factor(
      if_else(truth == current_class, current_class, "Other"),
      levels = c("Other", current_class)
    )
    estimate_binary <- factor(
      if_else(estimate == current_class, current_class, "Other"),
      levels = c("Other", current_class)
    )

    tibble(
      precision = precision_vec(
        truth_binary, estimate_binary, event_level = "second"
      ),
      recall = sens_vec(
        truth_binary, estimate_binary, event_level = "second"
      ),
      f1 = f_meas_vec(
        truth_binary, estimate_binary, event_level = "second"
      )
    )
  })

  tibble(
    accuracy = accuracy_vec(truth, estimate),
    macro_precision = mean(per_class$precision, na.rm = TRUE),
    macro_recall = mean(per_class$recall, na.rm = TRUE),
    macro_f1 = mean(per_class$f1, na.rm = TRUE),
    ari = adjusted_rand_index(truth, estimate)
  )
}
```

# 5. Zero-rule baseline

The zero-rule classifier predicts the most frequent training species for every
test specimen. Since the classes are balanced, its expected accuracy is about
one third.

```{r baseline}
majority_species <- train_data |>
  count(species, sort = TRUE) |>
  slice(1) |>
  pull(species) |>
  as.character()

baseline_class <- factor(
  rep(majority_species, nrow(test_data)),
  levels = levels(iris_data$species)
)

accuracy_vec(test_data$species, baseline_class)
```

# 6. Decision tree using original features

```{r feature-tree}
feature_tree_full <- rpart(
  species ~ sepal_length + sepal_width + petal_length + petal_width,
  data = train_data,
  method = "class",
  parms = list(split = "information"),
  control = rpart.control(cp = 0.001, minsplit = 10, xval = 10)
)

feature_tree_cp <- feature_tree_full$cptable[
  which.min(feature_tree_full$cptable[, "xerror"]),
  "CP"
]

feature_tree <- prune(feature_tree_full, cp = feature_tree_cp)

feature_tree_prediction <- predict(
  feature_tree,
  newdata = test_data,
  type = "class"
)
```

```{r plot-feature-tree, fig.height=7}
rpart.plot(
  feature_tree,
  type = 2,
  extra = 104,
  fallen.leaves = TRUE,
  box.palette = "GnBu",
  shadow.col = "gray",
  nn = TRUE
)
```

```{r feature-importance}
enframe(
  feature_tree$variable.importance,
  name = "measurement",
  value = "importance"
) |>
  mutate(relative_importance = importance / sum(importance)) |>
  arrange(desc(relative_importance))
```

# 7. Optimize K-means feature combinations

K-means is sensitive to scale, so every candidate feature set is standardized
using training statistics. We test all nonempty combinations of the four
features and candidate values of \(k\) from 2 to 6.

The selection criterion is **average silhouette width**:

- Values near 1 indicate compact, well-separated clusters.
- Values near 0 indicate overlapping clusters.
- Negative values suggest possible misassignment.

Species labels are not used in this search.

```{r feature-combination-search}
feature_combinations <- map(
  seq_along(feature_names),
  ~ combn(feature_names, .x, simplify = FALSE)
) |>
  flatten()

evaluate_kmeans_candidate <- function(features, k) {
  candidate_matrix <- train_data |>
    select(all_of(features)) |>
    scale()

  set.seed(123)
  fitted <- kmeans(candidate_matrix, centers = k, nstart = 50)
  silhouette_values <- silhouette(
    fitted$cluster,
    dist(candidate_matrix)
  )

  tibble(
    features = paste(features, collapse = " + "),
    feature_count = length(features),
    k = k,
    silhouette = mean(silhouette_values[, "sil_width"])
  )
}

kmeans_search <- map_dfr(feature_combinations, function(features) {
  map_dfr(2:6, ~ evaluate_kmeans_candidate(features, .x))
})

kmeans_search |>
  arrange(desc(silhouette)) |>
  slice_head(n = 15)
```

```{r optimal-feature-model}
best_feature_candidate <- kmeans_search |>
  slice_max(silhouette, n = 1, with_ties = FALSE)

best_feature_names <- str_split(
  best_feature_candidate$features,
  fixed(" + ")
)[[1]]

best_feature_k <- best_feature_candidate$k[[1]]

best_feature_candidate
```

```{r silhouette-search-chart}
kmeans_search |>
  group_by(features) |>
  slice_max(silhouette, n = 1, with_ties = FALSE) |>
  ungroup() |>
  slice_max(silhouette, n = 10) |>
  mutate(features = reorder(features, silhouette)) |>
  ggplot(aes(features, silhouette, fill = factor(k))) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Best K-means result for each leading feature combination",
    subtitle = "Selection uses training-data silhouette score, not species labels",
    x = "Feature combination",
    y = "Average silhouette width",
    fill = "Optimal k"
  ) +
  theme_minimal()
```

## Fit and apply the best feature-based K-means model

```{r best-feature-kmeans}
feature_scaled <- standardize_pair(
  train_data |> select(all_of(best_feature_names)),
  test_data |> select(all_of(best_feature_names))
)

set.seed(123)
best_feature_kmeans <- kmeans(
  feature_scaled$train,
  centers = best_feature_k,
  nstart = 100
)

feature_cluster_map <- learn_cluster_labels(
  best_feature_kmeans$cluster,
  train_data$species
)

feature_test_clusters <- nearest_cluster(
  feature_scaled$test,
  best_feature_kmeans$centers
)

feature_kmeans_prediction <- map_clusters_to_species(
  feature_test_clusters,
  feature_cluster_map,
  levels(iris_data$species)
)

feature_cluster_map
```

## Compare training clusters with actual classes

```{r feature-cluster-class-comparison}
table(
  Actual_species = train_data$species,
  Kmeans_cluster = best_feature_kmeans$cluster
)

adjusted_rand_index(
  train_data$species,
  best_feature_kmeans$cluster
)
```

The Adjusted Rand Index (ARI) compares two partitions without requiring cluster
numbers to match species names. A value near 1 indicates strong agreement; a
value near 0 indicates agreement similar to chance.

# 8. Principal component analysis

PCA transforms correlated measurements into uncorrelated components. It is fit
on standardized training measurements only. We retain the smallest number of
components whose cumulative explained variance reaches 90%.

```{r pca-fit}
pca_model <- prcomp(
  train_data |> select(all_of(feature_names)),
  center = TRUE,
  scale. = TRUE
)

pca_variance <- tibble(
  component = seq_along(pca_model$sdev),
  principal_component = paste0("PC", component),
  variance_explained = pca_model$sdev^2 / sum(pca_model$sdev^2)
) |>
  mutate(cumulative_variance = cumsum(variance_explained))

pca_variance
```

```{r pca-component-selection}
variance_threshold <- 0.90

optimal_pc_count <- which(
  pca_variance$cumulative_variance >= variance_threshold
)[1]

optimal_pc_names <- paste0("PC", seq_len(optimal_pc_count))

tibble(
  variance_threshold = percent(variance_threshold),
  retained_components = optimal_pc_count,
  retained_variance = percent(
    pca_variance$cumulative_variance[optimal_pc_count],
    accuracy = 0.1
  )
)
```

```{r scree-chart}
ggplot(pca_variance, aes(component, variance_explained)) +
  geom_col(fill = "#4C78A8") +
  geom_line(aes(y = cumulative_variance), color = "#E45756") +
  geom_point(aes(y = cumulative_variance), color = "#E45756") +
  geom_hline(
    yintercept = variance_threshold,
    linetype = "dashed",
    color = "#54A24B"
  ) +
  scale_x_continuous(breaks = pca_variance$component) +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  labs(
    title = "PCA explained variance",
    subtitle = "Bars: individual variance; red line: cumulative variance",
    x = "Principal component",
    y = "Share of standardized variance"
  ) +
  theme_minimal()
```

## Interpret component loadings

```{r pca-loadings}
as_tibble(
  pca_model$rotation[, optimal_pc_names, drop = FALSE],
  rownames = "measurement"
)
```

Large positive or negative loadings indicate measurements that contribute
strongly to a component. The sign indicates direction, not importance by itself.

## Transform training and test data

```{r pca-scores}
train_pc_scores <- predict(
  pca_model,
  newdata = train_data |> select(all_of(feature_names))
) |>
  as_tibble()

test_pc_scores <- predict(
  pca_model,
  newdata = test_data |> select(all_of(feature_names))
) |>
  as_tibble()

train_pc_data <- train_pc_scores |>
  select(all_of(optimal_pc_names)) |>
  mutate(species = train_data$species)

test_pc_data <- test_pc_scores |>
  select(all_of(optimal_pc_names)) |>
  mutate(species = test_data$species)
```

```{r pca-species-chart}
if (optimal_pc_count >= 2) {
  ggplot(
    train_pc_data,
    aes(PC1, PC2, color = species, shape = species)
  ) +
    geom_point(size = 3, alpha = 0.8) +
    labs(
      title = "Training specimens in principal-component space",
      color = "Species",
      shape = "Species"
    ) +
    theme_minimal()
}
```

# 9. Decision tree using optimal principal components

```{r pca-tree}
pca_tree_full <- rpart(
  species ~ .,
  data = train_pc_data,
  method = "class",
  parms = list(split = "information"),
  control = rpart.control(cp = 0.001, minsplit = 10, xval = 10)
)

pca_tree_cp <- pca_tree_full$cptable[
  which.min(pca_tree_full$cptable[, "xerror"]),
  "CP"
]

pca_tree <- prune(pca_tree_full, cp = pca_tree_cp)

pca_tree_prediction <- predict(
  pca_tree,
  newdata = test_pc_data,
  type = "class"
)
```

```{r plot-pca-tree, fig.height=7}
rpart.plot(
  pca_tree,
  type = 2,
  extra = 104,
  fallen.leaves = TRUE,
  box.palette = "GnBu",
  shadow.col = "gray",
  nn = TRUE
)
```

# 10. K-means using optimal principal components

The retained PCs already come from standardized inputs and are mutually
uncorrelated. We choose \(k\) using training-data silhouette score.

```{r pca-k-search}
pca_train_matrix <- train_pc_scores |>
  select(all_of(optimal_pc_names)) |>
  as.matrix()

pca_test_matrix <- test_pc_scores |>
  select(all_of(optimal_pc_names)) |>
  as.matrix()

pca_k_search <- map_dfr(2:6, function(k) {
  set.seed(123)
  fitted <- kmeans(pca_train_matrix, centers = k, nstart = 50)
  silhouette_values <- silhouette(
    fitted$cluster,
    dist(pca_train_matrix)
  )

  tibble(
    k = k,
    silhouette = mean(silhouette_values[, "sil_width"])
  )
})

pca_k_search
```

```{r best-pca-kmeans}
best_pca_k <- pca_k_search |>
  slice_max(silhouette, n = 1, with_ties = FALSE) |>
  pull(k)

set.seed(123)
pca_kmeans <- kmeans(
  pca_train_matrix,
  centers = best_pca_k,
  nstart = 100
)

pca_cluster_map <- learn_cluster_labels(
  pca_kmeans$cluster,
  train_data$species
)

pca_test_clusters <- nearest_cluster(
  pca_test_matrix,
  pca_kmeans$centers
)

pca_kmeans_prediction <- map_clusters_to_species(
  pca_test_clusters,
  pca_cluster_map,
  levels(iris_data$species)
)

tibble(
  retained_components = optimal_pc_count,
  optimal_k = best_pca_k,
  silhouette = max(pca_k_search$silhouette)
)
```

```{r pca-cluster-class-comparison}
table(
  Actual_species = train_data$species,
  PCA_Kmeans_cluster = pca_kmeans$cluster
)

adjusted_rand_index(
  train_data$species,
  pca_kmeans$cluster
)
```

# 11. Final comparison of four approaches

For a fair final comparison, all four approaches are evaluated on the same
untouched test specimens.

For K-means, training clusters are mapped to their majority species and test
specimens are assigned to the nearest centroid. Accuracy, precision, recall,
and F1 therefore measure recovery of the known species after mapping. ARI is
also reported because it evaluates partition agreement without depending on
label names.

```{r final-comparison}
comparison_results <- bind_rows(
  macro_classification_metrics(
    test_data$species,
    feature_tree_prediction
  ) |>
    mutate(model = "Decision tree: original features"),
  macro_classification_metrics(
    test_data$species,
    feature_kmeans_prediction
  ) |>
    mutate(model = "Best K-means: selected features"),
  macro_classification_metrics(
    test_data$species,
    pca_tree_prediction
  ) |>
    mutate(model = "Decision tree: selected PCs"),
  macro_classification_metrics(
    test_data$species,
    pca_kmeans_prediction
  ) |>
    mutate(model = "K-means: selected PCs")
) |>
  select(model, everything()) |>
  arrange(desc(accuracy), desc(macro_f1))

comparison_results |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

```{r final-comparison-chart, fig.height=7}
comparison_results |>
  pivot_longer(
    c(accuracy, macro_precision, macro_recall, macro_f1, ari),
    names_to = "metric",
    values_to = "estimate"
  ) |>
  ggplot(aes(reorder(model, estimate), estimate, fill = model)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ metric, ncol = 2) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Test performance of the four analytical approaches",
    x = NULL,
    y = "Metric value"
  ) +
  theme_minimal()
```

## Confusion matrices

```{r all-confusion-matrices}
prediction_sets <- list(
  "Decision tree: original features" = feature_tree_prediction,
  "Best K-means: selected features" = feature_kmeans_prediction,
  "Decision tree: selected PCs" = pca_tree_prediction,
  "K-means: selected PCs" = pca_kmeans_prediction
)

walk(names(prediction_sets), function(model_name) {
  cat("\n\n### ", model_name, "\n\n", sep = "")
  print(
    conf_mat(
      tibble(
        truth = test_data$species,
        estimate = prediction_sets[[model_name]]
      ),
      truth,
      estimate
    )
  )
})
```

## Model-selection summary

```{r selection-summary}
tibble(
  item = c(
    "Best raw-feature combination",
    "Best raw-feature K-means k",
    "Best raw-feature silhouette",
    "PCA variance threshold",
    "Retained principal components",
    "Best PCA K-means k",
    "Best PCA K-means silhouette"
  ),
  result = c(
    best_feature_candidate$features,
    as.character(best_feature_k),
    round(best_feature_candidate$silhouette, 3),
    percent(variance_threshold),
    paste(optimal_pc_names, collapse = " + "),
    as.character(best_pca_k),
    round(max(pca_k_search$silhouette), 3)
  )
)
```

# 12. Interpret the research findings

Use the outputs to answer:

1. Which original measurements most clearly separate the species?
2. Which feature combination and value of \(k\) maximize silhouette score?
3. Does the internally optimal \(k\) match the three known species? If not, what
   does that reveal about natural morphological groupings?
4. Which known species is recovered most cleanly by K-means?
5. How much standardized variance is retained by the selected PCs?
6. Which original measurements contribute most strongly to the retained PCs?
7. Does using PCA improve or reduce decision-tree performance?
8. Does using PCA improve cluster cohesion or agreement with known species?
9. Which of the four methods has the best test accuracy and macro F1 score?
10. Do silhouette and ARI select the same preferred clustering solution? Explain
    why an internally cohesive clustering can still disagree with known labels.

# 13. Limitations and responsible interpretation

- K-means finds geometric groups; it does not discover biological truth.
- Selecting the feature set by silhouette favors cluster geometry, not recovery
  of species labels.
- Mapping clusters to species is performed only to interpret and evaluate the
  clusters. It does not turn K-means into a supervised training algorithm.
- The Iris dataset is small, balanced, and historically used for teaching.
- A random split does not test generalization to another location, season, or
  measurement team.
- PCA maximizes retained variance, not classification accuracy or biological
  importance.
- Decision-tree splits and PCA loadings describe patterns, not biological causes.
- A single split can produce unstable rankings; repeated cross-validation or
  repeated train/test experiments should be used for stronger conclusions.

# 14. Optional learner exercises

1. Force \(k=3\) for every K-means candidate and compare the outcome with the
   silhouette-selected models.
2. Select the best feature combination using ARI instead of silhouette. Explain
   why this changes the problem from purely unsupervised selection.
3. Change the PCA threshold to 80% and 95%. How do the retained PCs and model
   rankings change?
4. Repeat the entire workflow across several random seeds and summarize the
   variability of each metric.
5. Compare the one-standard-error pruning rule with minimum cross-validation
   error for both decision trees.
6. Investigate misclassified specimens and propose additional measurements that
   could distinguish `versicolor` from `virginica`.

# Reproducibility information

```{r session-info}
sessionInfo()
```
