---
title: "Ames Housing: Linear Regression Lab"
subtitle: "Using property characteristics to support residential valuation"
author: "Analytics Solutions Bootcamp"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: flatly
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5
)

required_packages <- c(
  "modeldata", "tidyverse", "janitor", "rsample",
  "broom", "yardstick", "scales"
)

missing_packages <- required_packages[
  !required_packages %in% rownames(installed.packages())
]

if (length(missing_packages) > 0) {
  stop(
    "Install the missing packages first: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(modeldata)
library(tidyverse)
library(janitor)
library(rsample)
library(broom)
library(yardstick)
library(scales)

set.seed(123)
```

# Business scenario

A residential property company needs a consistent starting point for estimating
sale prices in Ames, Iowa. Agents currently rely heavily on comparable listings
and individual judgment. Management wants to understand how much price variation
can be explained by observable property characteristics and how large a typical
valuation error may be.

The model will support—not replace—professional appraisal. Historical sale price
is the numerical response variable, so this is a **regression problem**.

## Operational questions

This lab answers the following questions:

1. What is the typical sale price, and how widely do prices vary?
2. How much does sale price generally change as above-ground living area grows?
3. Is living area alone sufficient for initial valuation?
4. Does including age, basement area, garage capacity, bathrooms, and
   neighborhood materially improve estimates?
5. What is the typical dollar error for each model?
6. Which model better supports preliminary pricing and appraisal triage?
7. Which properties sell substantially above or below their estimated values
   and may require manual investigation?
8. What limitations should management consider before operational use?

> **Important:** Regression coefficients describe historical associations while
> holding included predictors constant. They do not establish that changing a
> property feature will cause its price to change by the coefficient amount.

# 1. Load and understand the data

```{r load-data}
data(ames, package = "modeldata")

ames_data <- ames |>
  clean_names()

glimpse(ames_data)
```

```{r data-quality}
tibble(
  properties = nrow(ames_data),
  variables = ncol(ames_data),
  duplicate_rows = sum(duplicated(ames_data)),
  missing_values = sum(is.na(ames_data))
)
```

For this introductory lab, we use a small, interpretable subset of variables.
This keeps the multiple-regression model aligned with the operational questions.

```{r select-variables}
housing_data <- ames_data |>
  transmute(
    sale_price,
    gr_liv_area,
    year_built,
    total_bsmt_sf,
    garage_cars,
    full_bath,
    neighborhood
  ) |>
  drop_na()

glimpse(housing_data)
```

| Variable | Operational meaning |
|---|---|
| `sale_price` | Historical property sale price in US dollars |
| `gr_liv_area` | Above-ground living area in square feet |
| `year_built` | Original construction year |
| `total_bsmt_sf` | Total basement area in square feet |
| `garage_cars` | Approximate garage capacity in cars |
| `full_bath` | Number of full bathrooms above ground |
| `neighborhood` | Physical neighborhood classification |

# 2. Explore the housing market

## Sale-price profile

```{r price-summary}
housing_data |>
  summarise(
    properties = n(),
    mean_price = mean(sale_price),
    median_price = median(sale_price),
    minimum_price = min(sale_price),
    maximum_price = max(sale_price),
    standard_deviation = sd(sale_price)
  )
```

```{r price-distribution}
price_mean <- mean(housing_data$sale_price)
price_median <- median(housing_data$sale_price)

ggplot(housing_data, aes(sale_price)) +
  geom_histogram(bins = 40, fill = "#4C78A8", color = "white") +
  geom_vline(
    xintercept = price_mean,
    color = "#E45756",
    linewidth = 1,
    linetype = "dashed"
  ) +
  geom_vline(
    xintercept = price_median,
    color = "#54A24B",
    linewidth = 1,
    linetype = "dotted"
  ) +
  scale_x_continuous(labels = dollar) +
  labs(
    title = "Distribution of residential sale prices",
    subtitle = "Dashed red: mean; dotted green: median",
    x = "Sale price",
    y = "Number of properties"
  ) +
  theme_minimal()
```

If the mean is much higher than the median, unusually expensive properties may
be pulling the average upward. That affects both operational reporting and
model errors.

## Price and living area

```{r price-area-relationship}
ggplot(housing_data, aes(gr_liv_area, sale_price)) +
  geom_point(alpha = 0.35, color = "#4C78A8") +
  geom_smooth(method = "lm", se = TRUE, color = "#E45756") +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Sale price and above-ground living area",
    subtitle = "The fitted line summarizes the average linear relationship",
    x = "Above-ground living area (sq ft)",
    y = "Sale price"
  ) +
  theme_minimal()
```

```{r area-price-correlation}
housing_data |>
  summarise(
    correlation = cor(gr_liv_area, sale_price),
    covariance = cov(gr_liv_area, sale_price)
  )
```

## Price differences across neighborhoods

```{r neighborhood-summary}
neighborhood_summary <- housing_data |>
  group_by(neighborhood) |>
  summarise(
    sales = n(),
    median_price = median(sale_price),
    mean_price = mean(sale_price),
    .groups = "drop"
  ) |>
  arrange(desc(median_price))

neighborhood_summary
```

```{r neighborhood-chart, fig.height=8}
ggplot(
  housing_data,
  aes(
    x = reorder(neighborhood, sale_price, FUN = median),
    y = sale_price
  )
) +
  geom_boxplot(fill = "#72B7B2", outlier.alpha = 0.35) +
  coord_flip() +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Sale-price distributions across neighborhoods",
    x = "Neighborhood",
    y = "Sale price"
  ) +
  theme_minimal()
```

### Operational discussion

- Would a company-wide average be a useful estimate for every property?
- Which market segments contain the highest-value properties?
- Do properties with similar living areas always sell for similar prices?
- Which omitted features might explain the remaining variation?

# 3. Prepare training and test data

We split the data before fitting either model. Stratification by `sale_price`
helps maintain a similar price distribution in both samples.

```{r split-data}
ames_split <- initial_split(
  housing_data,
  prop = 0.80,
  strata = sale_price
)

train_data <- training(ames_split)
test_data <- testing(ames_split)

bind_rows(
  train_data |>
    summarise(
      dataset = "Training",
      properties = n(),
      mean_price = mean(sale_price),
      median_price = median(sale_price)
    ),
  test_data |>
    summarise(
      dataset = "Test",
      properties = n(),
      mean_price = mean(sale_price),
      median_price = median(sale_price)
    )
)
```

# 4. Establish a baseline estimate

Before regression, suppose the company estimates every test property using the
mean training sale price. This gives a basic benchmark for evaluating whether
property characteristics add useful information.

```{r baseline}
training_mean_price <- mean(train_data$sale_price)

baseline_predictions <- test_data |>
  transmute(
    truth = sale_price,
    .pred = training_mean_price
  )

baseline_predictions |>
  summarise(
    model = "Training-mean baseline",
    mae = mae_vec(truth, .pred),
    mse = mse_vec(truth, .pred),
    rmse = rmse_vec(truth, .pred),
    rsq = rsq_vec(truth, .pred)
  )
```

# 5. Simple linear regression

The first model estimates sale price from above-ground living area alone:

\[
\widehat{SalePrice} = \beta_0 + \beta_1(GrLivArea)
\]

The fitted coefficients minimize the sum of squared residuals, where a residual
is the difference between an observed price and its fitted price.

```{r simple-model}
simple_model <- lm(
  sale_price ~ gr_liv_area,
  data = train_data
)

summary(simple_model)
```

## Interpret the coefficients

```{r simple-coefficients}
tidy(simple_model, conf.int = TRUE)
```

Interpretation guide:

- **Intercept:** estimated price when living area is zero. This is needed to
  position the line mathematically, but it is not operationally meaningful
  because a zero-square-foot home is outside the useful data range.
- **Living-area coefficient:** expected difference in sale price associated
  with one additional square foot, on average.
- Multiplying the coefficient by 100 gives a more practical estimate for an
  additional 100 square feet.

```{r simple-business-interpretation}
area_coefficient <- coef(simple_model)[["gr_liv_area"]]

tibble(
  additional_area_sq_ft = c(1, 100, 500),
  estimated_price_difference = area_coefficient * additional_area_sq_ft
)
```

This is an association, not a guaranteed return from constructing additional
space. Quality, location, layout, and other factors may differ.

## Generate test predictions

```{r simple-predictions}
simple_predictions <- test_data |>
  transmute(
    truth = sale_price,
    .pred = predict(simple_model, newdata = test_data),
    residual = truth - .pred
  )
```

# 6. Multiple linear regression

Living area does not capture the full valuation context. We extend the model:

\[
\widehat{SalePrice} = \beta_0 + \beta_1X_1 + \beta_2X_2 + \cdots + \beta_mX_m
\]

The multiple model estimates the association of each predictor with price while
holding the other included predictors constant.

```{r multiple-model}
multiple_model <- lm(
  sale_price ~ gr_liv_area + year_built + total_bsmt_sf +
    garage_cars + full_bath + neighborhood,
  data = train_data
)

summary(multiple_model)
```

## Interpret selected coefficients

```{r multiple-coefficients}
multiple_coefficients <- tidy(
  multiple_model,
  conf.int = TRUE
) |>
  arrange(p.value)

multiple_coefficients
```

Interpretation reminders:

- A numerical coefficient represents an expected price difference for a
  one-unit increase, holding the other included variables constant.
- A neighborhood coefficient is interpreted relative to the model's reference
  neighborhood.
- A small p-value indicates evidence of an association under the model's
  assumptions; it does not measure business importance or causality.
- Confidence intervals communicate uncertainty around each estimated
  coefficient.

```{r reference-neighborhood}
tibble(
  reference_neighborhood = levels(train_data$neighborhood)[1]
)
```

## Compare training model summaries

```{r model-fit-summary}
bind_rows(
  glance(simple_model) |>
    transmute(
      model = "Simple linear regression",
      r_squared = r.squared,
      adjusted_r_squared = adj.r.squared,
      residual_standard_error = sigma
    ),
  glance(multiple_model) |>
    transmute(
      model = "Multiple linear regression",
      r_squared = r.squared,
      adjusted_r_squared = adj.r.squared,
      residual_standard_error = sigma
    )
)
```

Adjusted \(R^2\) accounts for the number of predictors and is more useful than
training \(R^2\) alone when comparing models of different complexity. Final
model assessment should still use the untouched test data.

## Generate test predictions

```{r multiple-predictions}
multiple_predictions <- test_data |>
  transmute(
    truth = sale_price,
    .pred = predict(multiple_model, newdata = test_data),
    residual = truth - .pred
  )
```

# 7. Evaluate operational accuracy

The following metrics answer different operational questions:

- **MAE:** the typical absolute dollar difference between actual and estimated
  prices.
- **MSE:** average squared error; heavily penalizes large mistakes.
- **RMSE:** also penalizes large mistakes but remains in dollars.
- **MAPE:** average absolute percentage error, useful for relative accuracy.
- **R-squared:** share of test-set price variation captured by the estimates.

```{r evaluation-function}
regression_metrics <- function(predictions, model_name) {
  predictions |>
    summarise(
      model = model_name,
      mae = mae_vec(truth, .pred),
      mse = mse_vec(truth, .pred),
      rmse = rmse_vec(truth, .pred),
      mape = mean(abs((truth - .pred) / truth)) * 100,
      rsq = rsq_vec(truth, .pred)
    )
}

model_metrics <- bind_rows(
  regression_metrics(
    baseline_predictions,
    "Training-mean baseline"
  ),
  regression_metrics(
    simple_predictions,
    "Simple linear regression"
  ),
  regression_metrics(
    multiple_predictions,
    "Multiple linear regression"
  )
)

model_metrics |>
  mutate(
    across(c(mae, mse, rmse), ~ dollar(.x, accuracy = 1)),
    mape = percent(mape / 100, accuracy = 0.1),
    rsq = round(rsq, 3)
  )
```

## Quantify the improvement

```{r improvement-summary}
simple_mae <- model_metrics |>
  filter(model == "Simple linear regression") |>
  pull(mae)

multiple_mae <- model_metrics |>
  filter(model == "Multiple linear regression") |>
  pull(mae)

simple_rmse <- model_metrics |>
  filter(model == "Simple linear regression") |>
  pull(rmse)

multiple_rmse <- model_metrics |>
  filter(model == "Multiple linear regression") |>
  pull(rmse)

tibble(
  metric = c("MAE", "RMSE"),
  simple_model = c(simple_mae, simple_rmse),
  multiple_model = c(multiple_mae, multiple_rmse),
  reduction = c(
    1 - multiple_mae / simple_mae,
    1 - multiple_rmse / simple_rmse
  )
) |>
  mutate(
    across(c(simple_model, multiple_model), ~ dollar(.x, accuracy = 1)),
    reduction = percent(reduction, accuracy = 0.1)
  )
```

## Actual versus estimated prices

```{r actual-vs-predicted}
bind_rows(
  simple_predictions |>
    mutate(model = "Simple linear regression"),
  multiple_predictions |>
    mutate(model = "Multiple linear regression")
) |>
  ggplot(aes(truth, .pred)) +
  geom_point(alpha = 0.45, color = "#4C78A8") +
  geom_abline(slope = 1, intercept = 0, color = "#E45756", linetype = "dashed") +
  facet_wrap(~ model) +
  coord_equal() +
  scale_x_continuous(labels = dollar) +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Actual versus estimated sale prices",
    subtitle = "Points near the dashed line have smaller valuation errors",
    x = "Actual sale price",
    y = "Estimated sale price"
  ) +
  theme_minimal()
```

# 8. Diagnose the multiple regression model

## Residuals versus fitted values

```{r residual-plot}
augment(multiple_model) |>
  ggplot(aes(.fitted, .resid)) +
  geom_point(alpha = 0.35, color = "#4C78A8") +
  geom_hline(yintercept = 0, color = "#E45756", linetype = "dashed") +
  geom_smooth(se = FALSE, color = "#54A24B") +
  scale_x_continuous(labels = dollar) +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Residuals versus fitted prices",
    subtitle = "Systematic patterns suggest that linear-model assumptions may be incomplete",
    x = "Fitted sale price",
    y = "Residual: actual minus fitted price"
  ) +
  theme_minimal()
```

Operational interpretation:

- Residuals above zero represent underestimation.
- Residuals below zero represent overestimation.
- A widening spread at higher fitted prices suggests less consistent dollar
  accuracy for expensive properties.
- Curved patterns suggest nonlinear relationships or missing terms.

## Normal Q-Q plot

```{r qq-plot}
augment(multiple_model) |>
  ggplot(aes(sample = .std.resid)) +
  stat_qq(alpha = 0.4, color = "#4C78A8") +
  stat_qq_line(color = "#E45756") +
  labs(
    title = "Normal Q-Q plot of standardized residuals",
    x = "Theoretical quantiles",
    y = "Standardized residuals"
  ) +
  theme_minimal()
```

## Influential training observations

```{r influential-observations}
influence_summary <- augment(multiple_model) |>
  mutate(training_row = row_number()) |>
  arrange(desc(.cooksd)) |>
  select(
    training_row,
    sale_price,
    .fitted,
    .resid,
    .std.resid,
    .hat,
    .cooksd
  ) |>
  slice_head(n = 10)

influence_summary
```

Influential observations should be investigated, not automatically deleted.
They may be valid high-value properties, data-quality problems, or market
segments that require a different model.

# 9. Support valuation review

## Properties with the largest test errors

```{r largest-test-errors}
valuation_review <- test_data |>
  mutate(
    estimated_price = predict(multiple_model, newdata = test_data),
    residual = sale_price - estimated_price,
    absolute_error = abs(residual),
    error_direction = case_when(
      residual > 0 ~ "Model underestimated sale price",
      residual < 0 ~ "Model overestimated sale price",
      TRUE ~ "Exact estimate"
    )
  ) |>
  arrange(desc(absolute_error))

valuation_review |>
  select(
    sale_price,
    estimated_price,
    absolute_error,
    error_direction,
    gr_liv_area,
    year_built,
    neighborhood
  ) |>
  slice_head(n = 15)
```

These records form an **appraisal-review queue**, not proof that a transaction
was incorrectly priced. Large errors may point to omitted features, exceptional
property condition, unusual market circumstances, or data issues.

## Provide a valuation range

A prediction interval communicates uncertainty for an individual property's
sale price. The example below calculates 95% prediction intervals for a small
sample of test properties.

```{r prediction-intervals}
interval_sample <- test_data |>
  slice_head(n = 10)

prediction_intervals <- predict(
  multiple_model,
  newdata = interval_sample,
  interval = "prediction",
  level = 0.95
) |>
  as_tibble() |>
  rename(
    estimated_price = fit,
    lower_estimate = lwr,
    upper_estimate = upr
  )

bind_cols(
  interval_sample |>
    select(sale_price, gr_liv_area, year_built, neighborhood),
  prediction_intervals
)
```

Prediction intervals are generally more appropriate for an individual property
than confidence intervals for the average price. Their validity still depends
on the model assumptions and similarity to the training market.

# 10. Operational recommendation questions

Use the generated results to answer:

1. What price should management use as a naive baseline, and why is it limited?
2. How much price difference is associated with an additional 100 square feet
   in the simple model?
3. How much of the training price variation is explained by living area alone?
4. How much do MAE and RMSE improve after adding other property characteristics?
5. Is the multiple model accurate enough for preliminary estimates? Define an
   acceptable error with business stakeholders before answering.
6. Does the model perform consistently across the price range?
7. Which large-error properties should appraisers investigate first?
8. Should the model produce a point estimate, a valuation range, or both?
9. What controls should prevent the estimate from being treated as an automatic
   final appraisal?

# 11. Recommended operational workflow

1. Validate incoming property measurements and category values.
2. Generate a preliminary estimate and prediction interval.
3. Route large, uncertain, or out-of-range valuations to an appraiser.
4. Record the appraiser's adjustments and reasons.
5. Compare estimates with eventual sale prices using MAE, RMSE, MAPE, and bias.
6. Monitor error by neighborhood, price band, and property type.
7. Retrain or retire the model when market conditions or data definitions change.

# 12. Limitations

- Ames transactions represent a particular place and historical period.
- The selected predictors omit condition, remodeling, lot characteristics, and
  many other potentially important features available in the full dataset.
- Linear regression assumes additive, linear relationships unless terms are
  explicitly transformed or interacted.
- Neighborhood coefficients may absorb unmeasured local differences and should
  not be interpreted as causal neighborhood effects.
- A random split does not measure performance in a future housing market. A
  date-based validation is preferable before operational deployment.
- Extreme properties may require specialized appraisal rather than automated
  estimation.

# 13. Optional learner exercises

1. Fit a second simple regression using `year_built`. Is it more useful than
   living area alone?
2. Add `overall_qual` to the multiple model and evaluate the change in MAE and
   adjusted \(R^2\).
3. Fit a model using `log(sale_price)` and compare residual patterns.
4. Test an interaction between living area and neighborhood.
5. Calculate mean error—not absolute error—to determine whether the model is
   systematically overestimating or underestimating prices.
6. Compare MAE across low-, middle-, and high-price segments.
7. Write a short recommendation for valuation managers that separates model
   evidence, operational judgment, and limitations.

# Reproducibility information

```{r session-info}
sessionInfo()
```
