---
title: "Prescriptive Analytics: Employee Retention Planning"
subtitle: "From attrition risk to budget-constrained action"
author: "Analytics Solutions Bootcamp"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: flatly
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5
)

required_packages <- c(
  "modeldata", "tidyverse", "janitor", "rsample",
  "recipes", "lpSolve", "scales", "broom"
)

missing_packages <- required_packages[
  !required_packages %in% rownames(installed.packages())
]

if (length(missing_packages) > 0) {
  stop(
    "Install the missing packages first: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(modeldata)
library(tidyverse)
library(janitor)
library(rsample)
library(recipes)
library(lpSolve)
library(scales)
library(broom)

set.seed(123)
```

# Lab scenario

An organization has a limited annual budget for employee-retention
programs. HR cannot offer every intervention to every employee group, so
management must decide which programs to fund and where to deploy them.

This lab extends attrition classification into **prescriptive
analytics**:

-   Predictive question: Who may be at risk of leaving?
-   Prescriptive question: What should HR do with that information?

The objective is to maximize the expected financial benefit of retention
initiatives while respecting the available budget and HR's
implementation capacity.

> **Responsible-use boundary:** The dataset is synthetic and the
> intervention costs and effects in this lab are assumptions.
> Recommendations are made for employee segments, not as an
> employee-level watchlist. A real program requires privacy review,
> fairness testing, voluntary and supportive interventions, causal
> evidence, and human oversight. Never use attrition scores for
> dismissal, promotion denial, surveillance, or other punitive
> decisions.

## Operational questions

1.  Which workforce segments have the greatest expected turnover
    exposure?
2.  Which retention intervention offers the highest expected monetary
    value for each segment?
3.  Given a fixed budget, which segment–intervention combinations should
    HR fund?
4.  How much expected turnover cost could the recommended portfolio
    avoid?
5.  How does the recommendation change when the budget or intervention
    effectiveness changes?
6.  Which assumptions are most important to validate before
    implementation?

# 1. Frame the decision problem

Following the components of a decision problem:

| Component     | Definition in this lab                                                            |
|------------------------------------|------------------------------------|
| Objective     | Maximize expected net benefit from retention programs                             |
| Alternatives  | No action, flexible work, career development, training, or compensation review    |
| Constraints   | Annual budget and maximum number of programs HR can implement                     |
| Uncertainty   | Whether employees would leave and whether an intervention would prevent attrition |
| Decision unit | One intervention for an employee segment                                          |

This is a **decision under risk** because probabilities can be
estimated, but the outcomes are not certain.

# 2. Load and inspect the data

```{r load-data}
data(attrition, package = "modeldata")

attrition_data <- attrition |>
  clean_names() |>
  mutate(attrition = factor(attrition, levels = c("No", "Yes")))

glimpse(attrition_data)
```

```{r data-quality-check}
tibble(
  employees = nrow(attrition_data),
  variables = ncol(attrition_data),
  duplicate_rows = sum(duplicated(attrition_data)),
  missing_values = sum(is.na(attrition_data)),
  observed_attrition_rate = mean(attrition_data$attrition == "Yes")
)
```

```{r segment-overview}
attrition_data |>
  group_by(job_role, over_time) |>
  summarise(
    employees = n(),
    employees_who_left = sum(attrition == "Yes"),
    historical_attrition_rate = mean(attrition == "Yes"),
    median_monthly_income = median(monthly_income),
    .groups = "drop"
  ) |>
  arrange(desc(historical_attrition_rate))
```

# 3. Estimate attrition probabilities

The classification model is an input to the decision—not the final
output. We create a stratified training/test split and treat the test
set as a simulated future planning population whose outcomes are not
used to make decisions.

Protected attributes and administrative identifiers are excluded from
the prediction model. Protected attributes should still be retained
separately for fairness auditing in a real implementation.

```{r split-data}
attrition_split <- initial_split(
  attrition_data,
  prop = 0.80,
  strata = attrition
)

training_data <- training(attrition_split)
planning_data <- testing(attrition_split)

tibble(
  dataset = c("Training data", "Planning population"),
  employees = c(nrow(training_data), nrow(planning_data)),
  attrition_rate = c(
    mean(training_data$attrition == "Yes"),
    mean(planning_data$attrition == "Yes")
  )
)
```

```{r prepare-model-data}
excluded_variables <- c(
  "employee_number",
  "employee_count",
  "over18",
  "standard_hours",
  "gender"
)

attrition_recipe <- recipe(attrition ~ ., data = training_data) |>
  # step_rm(all_of(excluded_variables)) |>
  step_novel(all_nominal_predictors()) |>
  step_unknown(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors())

prepared_recipe <- prep(attrition_recipe, training = training_data)

training_baked <- bake(prepared_recipe, new_data = NULL)
planning_baked <- bake(prepared_recipe, new_data = planning_data)
```

```{r fit-risk-model}
risk_model <- glm(
  attrition ~ .,
  data = training_baked,
  family = binomial()
)

planning_scored <- planning_data |>
  mutate(
    predicted_attrition_probability = predict(
      risk_model,
      newdata = planning_baked,
      type = "response"
    )
  )

summary(planning_scored$predicted_attrition_probability)
```

The predicted probability is an estimate—not a fact and not a causal
explanation. It should be recalibrated and monitored when used on a
genuinely future population.

# 4. Translate risk into expected turnover exposure

For demonstration, assume that replacing an employee costs approximately
six months of the employee's monthly income. This includes recruiting,
onboarding, vacancy, and productivity costs. In practice, Finance and HR
should estimate this value by role and validate it using organizational
data.

For employee $i$:

$$
\text{Expected turnover exposure}_i =
P(\text{attrition}_i) \times \text{estimated turnover cost}_i
$$

```{r expected-exposure}
replacement_months <- 6

planning_scored <- planning_scored |>
  mutate(
    estimated_turnover_cost = monthly_income * replacement_months,
    expected_turnover_exposure =
      predicted_attrition_probability * estimated_turnover_cost
  )

planning_scored |>
  summarise(
    employees = n(),
    mean_predicted_risk = mean(predicted_attrition_probability),
    estimated_total_turnover_cost = sum(estimated_turnover_cost),
    expected_turnover_exposure = sum(expected_turnover_exposure)
  )
```

## Aggregate to decision segments

The lab uses `job_role` and `over_time` to form operationally
understandable segments. Very small groups should be combined or
suppressed in production to reduce privacy and statistical-stability
risks.

```{r create-segments}
segment_summary <- planning_scored |>
  group_by(job_role, over_time) |>
  summarise(
    employees = n(),
    mean_predicted_risk = mean(predicted_attrition_probability),
    expected_attrition_cases = sum(predicted_attrition_probability),
    total_expected_exposure = sum(expected_turnover_exposure),
    .groups = "drop"
  ) |>
  mutate(
    segment_id = row_number(),
    segment = paste(job_role, paste0("Overtime: ", over_time), sep = " | ")
  )

segment_summary |>
  arrange(desc(total_expected_exposure))
```

```{r exposure-chart}
segment_summary |>
  slice_max(total_expected_exposure, n = 12) |>
  ggplot(aes(
    x = reorder(segment, total_expected_exposure),
    y = total_expected_exposure
  )) +
  geom_col(fill = "#4C78A8") +
  coord_flip() +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Segments with the greatest expected turnover exposure",
    subtitle = "Exposure combines estimated attrition probability and replacement cost",
    x = NULL,
    y = "Expected turnover exposure"
  ) +
  theme_minimal()
```

# 5. Define intervention alternatives

The following values are hypothetical scenario assumptions.
`effectiveness` represents the assumed relative reduction in attrition
risk among participating employees.

```{r intervention-assumptions}
interventions <- tribble(
  ~intervention,          ~cost_per_employee, ~effectiveness,
  "Flexible work",                       500,           0.10,
  "Career development",                 1500,           0.22,
  "Training program",                   1200,           0.18,
  "Compensation review",                2500,           0.30
)

interventions
```

## Expected monetary value

For segment $s$ and intervention $j$:

$$
\text{Expected avoided cost}_{sj} =
\text{Expected exposure}_{s} \times \text{Effectiveness}_{j}
$$

$$
\text{Expected net benefit}_{sj} =
\text{Expected avoided cost}_{sj} - \text{Program cost}_{sj}
$$

```{r build-decision-options}
decision_options <- crossing(
  segment_summary,
  interventions
) |>
  mutate(
    option_id = row_number(),
    program_cost = employees * cost_per_employee,
    expected_avoided_cost = total_expected_exposure * effectiveness,
    expected_net_benefit = expected_avoided_cost - program_cost,
    expected_roi = expected_net_benefit / program_cost
  )

decision_options |>
  arrange(desc(expected_net_benefit)) |>
  select(
    segment, intervention, employees, program_cost,
    expected_avoided_cost, expected_net_benefit, expected_roi
  ) |>
  slice_head(n = 15)
```

Negative expected net benefit means that the intervention costs more
than its estimated avoided turnover cost under the current assumptions.
"No action" is implicitly available because the optimizer is not
required to select an option for every segment.

# 6. Optimize the retention portfolio

Suppose HR has:

-   an annual budget of \$150,000; and
-   capacity to implement programs in at most six employee segments.

The decision variable $x_{sj}$ is binary:

-   $x_{sj}=1$: fund intervention $j$ for segment $s$;
-   $x_{sj}=0$: do not fund it.

The model maximizes total expected net benefit subject to:

1.  Total program cost cannot exceed the budget.
2.  At most one intervention may be selected for each segment.
3.  At most six segment-level programs may be implemented.

```{r optimization-inputs}
annual_budget <- 150000
maximum_programs <- 6

number_of_options <- nrow(decision_options)
number_of_segments <- nrow(segment_summary)

budget_constraint <- decision_options$program_cost

segment_constraints <- outer(
  segment_summary$segment_id,
  decision_options$segment_id,
  FUN = function(segment, option_segment) {
    as.numeric(segment == option_segment)
  }
)

program_count_constraint <- rep(1, number_of_options)

constraint_matrix <- rbind(
  budget_constraint,
  segment_constraints,
  program_count_constraint
)

constraint_directions <- c(
  "<=",
  rep("<=", number_of_segments),
  "<="
)

constraint_limits <- c(
  annual_budget,
  rep(1, number_of_segments),
  maximum_programs
)
```

```{r solve-optimization}
optimization_result <- lp(
  direction = "max",
  objective.in = decision_options$expected_net_benefit,
  const.mat = constraint_matrix,
  const.dir = constraint_directions,
  const.rhs = constraint_limits,
  all.bin = TRUE
)

if (optimization_result$status != 0) {
  stop("The optimization model did not find an optimal solution.")
}

recommended_portfolio <- decision_options |>
  mutate(selected = optimization_result$solution > 0.5) |>
  filter(selected) |>
  arrange(desc(expected_net_benefit))

recommended_portfolio |>
  select(
    segment, intervention, employees, mean_predicted_risk,
    program_cost, expected_avoided_cost,
    expected_net_benefit, expected_roi
  )
```

```{r portfolio-summary}
portfolio_summary <- recommended_portfolio |>
  summarise(
    selected_programs = n(),
    employees_covered = sum(employees),
    total_program_cost = sum(program_cost),
    unused_budget = annual_budget - total_program_cost,
    expected_avoided_cost = sum(expected_avoided_cost),
    expected_net_benefit = sum(expected_net_benefit),
    portfolio_roi = expected_net_benefit / total_program_cost
  )

portfolio_summary
```

## Compare with a simple ranking rule

A common heuristic is to fund options from highest to lowest predicted
risk. That approach ignores differences in employee count, turnover
cost, program cost, and intervention effectiveness. The optimized
recommendation explicitly balances these trade-offs.

```{r compare-risk-ranking}
risk_priority <- decision_options |>
  group_by(segment_id) |>
  slice_max(effectiveness, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(desc(mean_predicted_risk)) |>
  mutate(
    cumulative_cost = cumsum(program_cost),
    rank = row_number()
  ) |>
  filter(
    cumulative_cost <= annual_budget,
    rank <= maximum_programs
  )

bind_rows(
  recommended_portfolio |>
    summarise(
      approach = "Optimized portfolio",
      programs = n(),
      cost = sum(program_cost),
      expected_avoided_cost = sum(expected_avoided_cost),
      expected_net_benefit = sum(expected_net_benefit)
    ),
  risk_priority |>
    summarise(
      approach = "Highest-risk-first heuristic",
      programs = n(),
      cost = sum(program_cost),
      expected_avoided_cost = sum(expected_avoided_cost),
      expected_net_benefit = sum(expected_net_benefit)
    )
)
```

# 7. Budget scenario analysis

Optimization is most useful when decision makers can ask, "What changes
if the constraint changes?" The function below resolves the portfolio
for different budget levels.

```{r optimization-function}
optimize_portfolio <- function(options, budget, max_programs = 6) {
  n_options <- nrow(options)
  segment_ids <- sort(unique(options$segment_id))

  segment_matrix <- outer(
  segment_ids,
  options$segment_id,
  FUN = function(segment, option_segment) {
    as.numeric(segment == option_segment)
  }
)

  constraint_matrix <- rbind(
    options$program_cost,
    segment_matrix,
    rep(1, n_options)
  )

  result <- lp(
    direction = "max",
    objective.in = options$expected_net_benefit,
    const.mat = constraint_matrix,
    const.dir = c("<=", rep("<=", length(segment_ids)), "<="),
    const.rhs = c(budget, rep(1, length(segment_ids)), max_programs),
    all.bin = TRUE
  )

  if (result$status != 0) {
    return(tibble(
      budget = budget,
      selected_programs = NA_integer_,
      total_cost = NA_real_,
      expected_avoided_cost = NA_real_,
      expected_net_benefit = NA_real_
    ))
  }

  selected <- options[result$solution > 0.5, ]

  tibble(
    budget = budget,
    selected_programs = nrow(selected),
    total_cost = sum(selected$program_cost),
    expected_avoided_cost = sum(selected$expected_avoided_cost),
    expected_net_benefit = sum(selected$expected_net_benefit)
  )
}
```

```{r budget-scenarios}
budget_results <- map_dfr(
  seq(50000, 300000, by = 25000),
  ~ optimize_portfolio(decision_options, budget = .x)
)

budget_results
```

```{r budget-scenario-chart}
budget_results |>
  ggplot(aes(budget, expected_net_benefit)) +
  geom_line(color = "#4C78A8", linewidth = 1) +
  geom_point(color = "#4C78A8", size = 2.5) +
  scale_x_continuous(labels = dollar) +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Expected net benefit at different HR budgets",
    subtitle = "A flat section indicates that additional budget does not improve the current solution",
    x = "Available retention budget",
    y = "Maximum expected net benefit"
  ) +
  theme_minimal()
```

# 8. Effectiveness sensitivity analysis

Intervention effectiveness is uncertain. Test pessimistic, base, and
optimistic assumptions by applying a multiplier to every intervention's
effectiveness.

```{r effectiveness-scenarios}
effectiveness_scenarios <- tribble(
  ~scenario,      ~effectiveness_multiplier,
  "Pessimistic",                      0.60,
  "Base",                             1.00,
  "Optimistic",                       1.30
)

effectiveness_results <- effectiveness_scenarios |>
  mutate(
    options = map(effectiveness_multiplier, function(multiplier) {
      decision_options |>
        mutate(
          scenario_effectiveness = pmin(effectiveness * multiplier, 1),
          expected_avoided_cost =
            total_expected_exposure * scenario_effectiveness,
          expected_net_benefit = expected_avoided_cost - program_cost
        )
    }),
    result = map(
      options,
      ~ optimize_portfolio(
        .x,
        budget = annual_budget,
        max_programs = maximum_programs
      )
    )
  ) |>
  select(scenario, effectiveness_multiplier, result) |>
  unnest(result)

effectiveness_results
```

## Interpretation questions

-   Does the recommended portfolio remain valuable under pessimistic
    assumptions?
-   Which recommendation changes first when effectiveness declines?
-   At what budget level do expected benefits begin to flatten?
-   Is unused budget evidence of an error, or are no additional options
    valuable under the model assumptions?
-   Which assumptions should HR test through a small pilot before
    scaling?

# 9. Fairness, privacy, and governance checks

Optimization can make a mathematically efficient recommendation that is
still unfair, infeasible, or harmful. Review the selected segments
before acting.

```{r governance-review}
governance_review <- recommended_portfolio |>
  transmute(
    segment,
    intervention,
    employees,
    mean_predicted_risk,
    expected_net_benefit,
    review_small_group = employees < 10,
    review_high_risk_label = mean_predicted_risk >= 0.50
  )

governance_review
```

Required controls for a real deployment include:

1.  **Data minimization:** Use only information necessary for supportive
    action.
2.  **Access control:** Restrict employee-level data and model outputs.
3.  **Fairness audit:** Examine whether error rates and benefits differ
    across protected groups, even when protected attributes are excluded
    as predictors.
4.  **Causal validation:** Do not assume observational associations
    prove that an intervention will reduce attrition.
5.  **Human review:** Managers and HR should assess feasibility and
    employee welfare before implementation.
6.  **Voluntary participation:** Retention programs should benefit
    employees and avoid coercion or stigma.
7.  **Monitoring:** Compare projected and realized costs, participation,
    employee outcomes, and attrition after deployment.

# 10. Recommended operational workflow

1.  Validate the attrition model on recent, representative data.
2.  Estimate actual turnover costs with Finance and operating teams.
3.  Estimate intervention effects using historical evaluations or a
    pilot.
4.  Define budget, capacity, policy, and fairness constraints with
    stakeholders.
5.  Generate the optimized portfolio.
6.  Conduct human, privacy, fairness, and feasibility reviews.
7.  Pilot the intervention with employee consent.
8.  Measure outcomes and update the assumptions before scaling.

# 11. Conclusions

Complete these statements after running the lab:

-   The planning population contains `r nrow(planning_scored)` employees
    across `r nrow(segment_summary)` segments.
-   Under a budget of `r dollar(annual_budget)`, the optimization
    selects `r nrow(recommended_portfolio)` segment-level programs.
-   The recommended portfolio costs
    `r dollar(sum(recommended_portfolio$program_cost))` and has an
    expected net benefit of
    `r dollar(sum(recommended_portfolio$expected_net_benefit))`.
-   These are model-based expectations, not guaranteed savings.
-   The recommendation is valid only to the extent that the risk,
    turnover-cost, and intervention-effect assumptions are credible.

# 12. Learner exercises

1.  Change the annual budget to \$75,000 and \$250,000. Which
    recommendations change, and why?
2.  Add a constraint that at least one non-overtime segment must receive
    a program. Explain the business justification and trade-off.
3.  Allow no more than two compensation-review programs. How does this
    affect expected net benefit?
4.  Replace six months of income with three, nine, and twelve months
    when estimating turnover cost. Compare the solutions.
5.  Create an intervention designed specifically for overtime-related
    workload. Define a defensible cost and effectiveness range.
6.  Compare the optimized portfolio with a highest-risk-first and a
    highest-ROI-first heuristic.
7.  Design a randomized pilot that could estimate intervention
    effectiveness more credibly.
8.  Identify sensitive variables that should not be used for
    intervention targeting, and explain why.

# 13. Key limitations

-   The attrition data are synthetic and may not represent a real
    organization.
-   The test set is used as a simulated future planning population.
-   Intervention costs, effectiveness, and turnover costs are
    assumptions.
-   Logistic regression estimates association, not the causal effect of
    an intervention.
-   Segment-level treatment assumes employees within a segment receive
    the same program and benefit equally.
-   Expected monetary value summarizes uncertainty but does not
    eliminate it.
-   Financial benefit is only one objective; employee well-being,
    equity, legal requirements, and organizational feasibility must also
    guide decisions.
