---
title: "Exploratory Data Analysis and Basic Statistical Tests with Diamonds"
author: "Josh Valdeleon"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

# install.packages("tidyverse")
# install.packages("plotly")

library(tidyverse)
library(plotly)
```

# Learning Objectives

By the end of this demo, learners should be able to:

- Inspect and profile a dataset before analysis.
- Perform univariate, bivariate, and multivariate exploratory analysis.
- Visualize relationships between numerical and categorical variables.
- Formulate null and alternative hypotheses.
- Perform and interpret basic statistical tests.
- Distinguish statistical significance from practical significance.
- Recognize how confounding variables can affect conclusions.

# Dataset

For this exercise, we will use the `diamonds` dataset included in `ggplot2`.

The dataset contains information about diamond prices and characteristics such as carat, cut, color, and clarity.

```{r load-data}
data("diamonds", package = "ggplot2")

diamonds
```

# 1. Data Inspection

Before performing any analysis, inspect the structure of the dataset.

```{r inspect-data}
glimpse(diamonds)
```

Check the dimensions.

```{r dimensions}
dim(diamonds)
```

View basic descriptive statistics.

```{r summary}
summary(diamonds)
```

Check for missing values.

```{r missing-values}
diamonds |>
  summarise(
    across(
      everything(),
      ~ sum(is.na(.))
    )
  )
```

# 2. Univariate Exploratory Analysis

## 2.1 Distribution of Diamond Price

```{r price-histogram}
mean_price <- mean(diamonds$price)
median_price <- median(diamonds$price)

diamonds |>
  ggplot(aes(x = price)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = mean_price, color='red') + 
  geom_vline(xintercept = median_price, color='blue') + 
  labs(
    title = "Distribution of Diamond Prices",
    subtitle = "Red = Mean | Blue = Median",
    x = "Price",
    y = "Count"
  ) +
  theme_minimal()
```

Calculate descriptive statistics for price.

```{r price-summary}
diamonds |>
  summarise(
    n = n(),
    mean_price = mean(price),
    median_price = median(price),
    min_price = min(price),
    max_price = max(price),
    sd_price = sd(price)
  )
```

### Discussion Questions

- Is the price distribution symmetric or skewed?
- Is the mean price higher or lower than the median?
- Are there possible outliers?

## 2.2 Distribution of Carat

```{r carat-histogram}
diamonds |>
  ggplot(aes(x = carat)) +
  geom_histogram(bins = 50) +
  labs(
    title = "Distribution of Diamond Carat",
    x = "Carat",
    y = "Count"
  ) +
  theme_minimal()
```

## 2.3 Distribution of Diamond Cut

```{r cut-counts}
diamonds |>
  count(cut) |>
  arrange(desc(n))
```

```{r cut-bar-chart}
diamonds |>
  ggplot(aes(x = cut)) +
  geom_bar() +
  labs(
    title = "Number of Diamonds by Cut",
    x = "Cut",
    y = "Count"
  ) +
  theme_minimal()
```

```{r cut-bar-chart-mean}
diamonds |>
  ggplot(aes(x = cut, y = price)) +
  stat_summary(fun = mean, geom = 'bar') + 
  labs(
    title = "Average price of diamonds by Cut",
    x = "Cut",
    y = "Average Price"
  ) +
  theme_minimal()
```

# 3. Bivariate Exploratory Analysis

## 3.1 Carat and Price

**Question:** Do larger diamonds tend to have higher prices?

```{r carat-price-scatter}
diamonds |>
  ggplot(aes(x = carat, y = price)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm") +
  labs(
    title = "Diamond Price vs. Carat",
    x = "Carat",
    y = "Price"
  ) +
  theme_minimal()
```

### Discussion Questions

- What direction does the relationship appear to have?
- Does the relationship look linear?
- Does the variability in price change as carat increases?

## 3.2 Cut and Price

```{r cut-price-boxplot}
diamonds |>
  ggplot(aes(x = cut, y = price)) +
  geom_boxplot() +
  labs(
    title = "Diamond Price by Cut",
    x = "Cut",
    y = "Price"
  ) +
  theme_minimal()
```

Summarize price by cut.

```{r cut-price-summary}
diamonds |>
  group_by(cut) |>
  summarise(
    n = n(),
    mean_price = mean(price),
    median_price = median(price),
    mean_carat = mean(carat),
    .groups = "drop"
  )
```

### Discussion Question

A higher-quality cut might be expected to have a higher price.

Does the raw data support that expectation?

# 4. Multivariate Exploratory Analysis

Explore the correlation matrix of all numeric variables. 
```{r corr-matrix}
cor_matrix <- diamonds |>
  select(where(is.numeric)) |>
  cor()

cor_df <- cor_matrix |>
  as.data.frame() |>
  rownames_to_column("variable1") |>
  pivot_longer(
    -variable1,
    names_to = "variable2",
    values_to = "correlation"
  )

cor_df |>
  ggplot(
    aes(
      x = variable1,
      y = variable2,
      fill = correlation
    )
  ) +
  geom_tile() +
  geom_text(
    aes(label = round(correlation, 2))
  ) +
  labs(
    title = "Correlation Matrix",
    x = NULL,
    y = NULL,
    fill = "Correlation"
  ) +
  theme_minimal()
```

Explore the relationship between carat and price while also considering diamond cut.

```{r multivariate-scatter}
diamonds |>
  ggplot(
    aes(
      x = carat,
      y = price,
      color = cut
    )
  ) +
  geom_point(alpha = 0.3) +
  labs(
    title = "Diamond Price vs. Carat by Cut",
    x = "Carat",
    y = "Price",
    color = "Cut"
  ) +
  theme_minimal()
```

This visualization illustrates why analysts should be cautious when interpreting relationships from only one or two variables.

```{r 5d-chart}
diamonds |>
  slice_sample(n = 3000) |>
  plot_ly(
    x = ~x,
    y = ~y,
    z = ~z,
    color = ~cut,
    size = ~price,
    type = "scatter3d",
    mode = "markers",
    marker = list(
      opacity = 0.5
    )
  ) |>
  layout(
    title = "Diamond Dimensions by Cut and Price",
    scene = list(
      xaxis = list(title = "Length (x)"),
      yaxis = list(title = "Width (y)"),
      zaxis = list(title = "Depth (z)")
    )
  )
```

# 5. Statistical Test 1: Pearson Correlation

## Research Question

**Is diamond carat significantly associated with diamond price?**

### Hypotheses

- **H0:** There is no linear correlation between carat and price.
- **H1:** There is a linear correlation between carat and price.

First, calculate the correlation coefficient.

```{r correlation}
cor(
  diamonds$carat,
  diamonds$price
)
```

Perform a Pearson correlation test.

```{r correlation-test}
cor.test(
  diamonds$carat,
  diamonds$price,
  method = "pearson"
)
```

### Interpretation Guide

Focus on:

- correlation coefficient (`r`)
- direction of the relationship
- strength of the relationship
- p-value
- confidence interval

If the p-value is below the chosen significance level, such as 0.05, reject the null hypothesis.

However, remember:

> Statistical association does not automatically imply causation.

# 6. Statistical Test 2: Independent Samples t-test

## Research Question

**Do Ideal-cut and Premium-cut diamonds have different average prices?**

Create a subset containing only the two groups.

```{r ttest-data}
diamonds_two_cuts <- diamonds |>
  filter(cut %in% c("Ideal", "Premium"))

diamonds_two_cuts |>
  count(cut)
```

Compare their descriptive statistics.

```{r ttest-summary}
diamonds_two_cuts |>
  group_by(cut) |>
  summarise(
    n = n(),
    mean_price = mean(price),
    median_price = median(price),
    sd_price = sd(price),
    .groups = "drop"
  )
```

### Hypotheses

- **H0:** Mean price of Ideal diamonds = Mean price of Premium diamonds.
- **H1:** Mean price of Ideal diamonds != Mean price of Premium diamonds.

Perform the test.

```{r ttest}
t.test(
  price ~ cut,
  data = diamonds_two_cuts
)
```

### Interpretation Guide

If the p-value is below 0.05, there is evidence that the two groups have different mean prices.

Because the `diamonds` dataset is large, small differences can become statistically significant.

This means we should also ask:

> Is the difference large enough to matter in practice?

# 7. Statistical Test 3: One-Way ANOVA

## Research Question

**Does average diamond price differ across cut categories?**

### Hypotheses

- **H0:** All cut categories have the same population mean price.
- **H1:** At least one cut category has a different population mean price.

Fit the ANOVA model.

```{r anova}
price_cut_anova <- aov(
  price ~ cut,
  data = diamonds
)

summary(price_cut_anova)
```

### Interpretation

A statistically significant ANOVA tells us that at least one group differs.

It does **not** tell us exactly which groups differ.

# 8. Post-Hoc Analysis: Tukey HSD

If the ANOVA is statistically significant, perform pairwise comparisons.

```{r tukey}
TukeyHSD(price_cut_anova)
```

### Discussion Questions

- Which pairs of cut categories have statistically significant differences?
- Which comparisons show the largest differences?
- Are these differences necessarily caused by cut quality alone?

# 9. Statistical Test 4: Chi-Square Test of Independence

## Research Question

**Is diamond cut associated with diamond color?**

Both variables are categorical.

Create a contingency table.

```{r contingency-table}
cut_color_table <- table(
  diamonds$cut,
  diamonds$color
)

cut_color_table
```

### Hypotheses

- **H0:** Diamond cut and color are independent.
- **H1:** Diamond cut and color are associated.

Perform the chi-square test.

```{r chi-square}
chisq.test(cut_color_table)
```

Visualize the proportions.

```{r cut-color-chart}
diamonds |>
  ggplot(aes(x = cut, fill = color)) +
  geom_bar(position = "fill") +
  labs(
    title = "Distribution of Diamond Color within Each Cut",
    x = "Cut",
    y = "Proportion",
    fill = "Color"
  ) +
  theme_minimal()
```

### Interpretation Guide

A statistically significant chi-square result suggests that cut and color are not independent.

However, statistical association does not by itself explain why the variables are related.

# 10. Investigating a Surprising Finding

Return to the earlier question:

> Do better-cut diamonds cost more?

Inspect the raw average price and carat by cut.

```{r surprising-finding}
diamonds |>
  group_by(cut) |>
  summarise(
    n = n(),
    mean_price = mean(price),
    median_price = median(price),
    mean_carat = mean(carat),
    .groups = "drop"
  )
```

The raw average prices may not follow the simple pattern we might initially expect.

Why?

One possible explanation is **carat**.

Diamond price is strongly related to diamond size, and the distribution of carat differs across cut categories.

# 11. Controlling for Carat with Regression

We can examine the relationship between cut and price while controlling for carat using a linear regression model.

```{r regression}
price_model <- lm(
  price ~ carat + cut,
  data = diamonds
)

summary(price_model)
```

Compare this with a model using only cut.

```{r cut-only-model}
cut_only_model <- lm(
  price ~ cut,
  data = diamonds
)

summary(cut_only_model)
```

### Discussion Questions

- How does the interpretation of cut change after including carat?
- Why might the raw average price by cut be misleading?
- What other variables might affect diamond price?
- Does controlling for carat prove that cut causes changes in price?

# 12. Key Takeaways

Exploratory analysis and statistical testing answer related but different questions.

EDA helps us **discover patterns and generate questions**.

Statistical tests help us evaluate whether observed differences or relationships are likely to be consistent with a population-level relationship.

In this exercise, we used:

1. **Pearson correlation** for two numerical variables.
2. **Independent samples t-test** for comparing two groups.
3. **One-way ANOVA** for comparing more than two groups.
4. **Tukey HSD** for post-hoc comparisons.
5. **Chi-square test** for relationships between categorical variables.
6. **Linear regression** to examine relationships while considering another variable.

Most importantly:

> A statistically significant result is not automatically an important result, and an observed association does not automatically imply causation.

# 13. Challenge Questions

Try answering the following independently.

1. Is `depth` significantly correlated with `price`?
2. Do `Ideal` and `Premium` diamonds have different average carat sizes?
3. Does average carat differ across clarity categories?
4. Are diamond `cut` and `clarity` associated?
5. Does the relationship between carat and price differ across cut categories?
6. Build a regression model using `carat`, `cut`, `color`, and `clarity` to explain diamond price.
